Minimax-optimal Laplacian Eigenmaps regression over Sobolev Spaces with Neighborhood Graphs

### 2021-09-04 manuscripts/laplacian_eigenmaps/main_text.tex
Put Laplacian Eigenmaps paper into AOS format.

\documentclass[aos]{imsart}

%% Packages
\RequirePackage{amsthm,amsmath,amsfonts,amssymb}
% \RequirePackage[numbers]{natbib}
\RequirePackage[authoryear]{natbib}
\RequirePackage[colorlinks,citecolor=blue,urlcolor=blue]{hyperref}
\RequirePackage{graphicx}
\usepackage{mathtools, enumitem}
\usepackage{multirow}

\startlocaldefs
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%                                          %%
%% For Axiom, Claim, Corollary, Hypothesis, %%
%% Lemma, Theorem, Proposition              %%
%% use \theoremstyle{plain}                 %%
%%                                          %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\theoremstyle{plain}
\newtheorem{theorem}{Theorem}
\newtheorem{lemma}{Lemma}
\newtheorem{corollary}{Corollary}
\newtheorem{proposition}{Proposition}
\theoremstyle{definition}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%                                          %%
%% For Assumption, Definition, Example,     %%
%% Notation, Property, Remark, Fact         %%
%% use \theoremstyle{remark}                %%
%%                                          %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\theoremstyle{remark}
\newtheorem{remark}{Remark}
\newtheorem{definition}{Definition}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Please put your definitions here:        %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%% Begin Ryan's definitions

% Widebar
\makeatletter
\newcommand*\rel@kern[1]{\kern#1\dimexpr\macc@kerna}
\newcommand*\widebar[1]{%
	\begingroup
	\def\mathaccent##1##2{%
		\rel@kern{0.8}%
		\overline{\rel@kern{-0.8}\macc@nucleus\rel@kern{0.2}}%
		\rel@kern{-0.2}%
	}%
	\macc@depth\@ne
	\let\math@bgroup\@empty \let\math@egroup\macc@set@skewchar
	\mathsurround\z@ \frozen@everymath{\mathgroup\macc@group\relax}%
	\macc@set@skewchar\relax
	\let\mathaccentV\macc@nested@a
	\macc@nested@a\relax111{#1}%
	\endgroup
}
\makeatother

% Min and max
\newcommand{\argmin}{\mathop{\mathrm{argmin}}}
\newcommand{\argmax}{\mathop{\mathrm{argmax}}}
\newcommand{\minimize}{\mathop{\mathrm{minimize}}}
\newcommand{\st}{\mathop{\mathrm{subject\,\,to}}}

% Shortcuts
\def\R{\mathbb{R}}

%%% End Ryan's definitions

%%% Begin Alden's additions
\newcommand{\Ebb}{\mathbb{E}}
\newcommand{\Pbb}{\mathbb{P}}
\newcommand{\dotp}[2]{\langle #1, #2 \rangle}
\newcommand{\wt}[1]{\widetilde{#1}}
\newcommand{\wh}[1]{\widehat{#1}}
\newcommand{\mc}[1]{\mathcal{#1}}
\newcommand{\Reals}{\mathbb{R}} % Same thing as Ryan's \R
\newcommand{\Rd}{\Reals^d}
\newcommand{\wb}[1]{\widebar{#1}}
\newcommand{\floor}[1]{\left\lfloor #1 \right\rfloor}
\newcommand{\Var}{\mathrm{Var}}
\newcommand{\Cov}{\mathrm{Cov}}
\newcommand{\1}{\mathbf{1}}
\newcommand{\bj}{{\bf j}}
\newcommand{\restr}[2]{\ensuremath{\left.#1\right|_{#2}}}

\DeclareFontFamily{U}{mathx}{\hyphenchar\font45}
\DeclareFontShape{U}{mathx}{m}{n}{<-> mathx10}{}
\DeclareSymbolFont{mathx}{U}{mathx}{m}{n}
\DeclareMathAccent{\wc}{0}{mathx}{"71}
%%% End Alden's Additions

\endlocaldefs

\begin{document}

\begin{frontmatter}

\title{Minimax-optimal Regression over Sobolev Spaces via Laplacian Eigenmaps with Neighborhood Graphs}
\runtitle{Minimax-optimal Laplacian Eigenmaps Regression}
%\thankstext{T1}{A sample of additional note to the title.}

\begin{aug}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Only one address is permitted per author. %%
%% Only division, organization and e-mail is %%
%% included in the address.                  %%
%% Additional information can be included in %%
%% the Acknowledgments section if necessary. %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\author[A]{\fnms{Alden} \snm{Green}\ead[label=e1]{}},
\author[A]{\fnms{Sivaraman} \snm{Balakrishnan}\ead[label=e1]{}}
\and
\author[A]{\fnms{Ryan J.} \snm{Tibshirani}\ead[label=e1]{}}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Addresses                                %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\address[A]{Carnegie Mellon University, \printead{e1}}
\end{aug}

\begin{abstract}
In this paper we study the statistical properties of nonparametric regression methods based on \emph{Laplacian eigenmaps}. These are spectral projection methods, which involve projecting a vector of observed responses onto a subspace spanned by certain eigenvectors of a neighborhood graph Laplacian. We show that in many cases these methods achieve minimax rates of convergence over continuum function classes. Specifically, we consider the problem of random design regression over Sobolev spaces $H^s(\mc{X})$---where $\mc{X} \subseteq \Rd$ is the support of the design density $p$---and show that under sufficient smoothness conditions on $p$, methods based on Laplacian eigenmaps achieve the optimal rates for both estimation (where the optimal rate is known to be $n^{-2s/(2s + d)}$) and goodness-of-fit testing ($n^{-4s/(4s + d)}$). We also show that Laplacian eigenmaps is \emph{manifold adaptive}: that is, we consider the situation where $\mc{X}$ is a manifold of small intrinsic dimension $m$, and give upper bounds establishing that Laplacian eigenmaps achieves the faster minimax estimation ($n^{-2s/(2s + m)}$) and testing ($n^{-4s/(4s + m)}$) rates of convergence. Finally, since the Laplacian eigenmaps estimator is defined only at observed design points, we propose an out-of-sample extension based on kernel smoothing, and prove that it achieves optimal out-of-sample error. We support these theoretical results with empirical evidence.
\end{abstract}

\begin{keyword}[class=MSC]
\kwd{62G08, 62G10, 68R10}
\end{keyword}

\begin{keyword}
\kwd{Laplacian eigenmaps, nonparametric regression, Sobolev space, minimax convergence rate}
\end{keyword}

\end{frontmatter}
% Main text
\section{Introduction}
\label{sec:introduction}
Two fundamental statistical goals are to learn and leverage structure of an unknown distribution on the basis of samples. Neighborhood graph Laplacians, and their associated spectra, have proven to be powerful and general tools for both purposes. They have been successfully used for various statistical tasks such as (spectral) clustering, manifold learning, level-set estimation, semi-supervised learning, etc.

@@

More specifically, we consider the worst-case risk of Laplacian eigenmaps over normed balls in Sobolev spaces. Roughly speaking, for a domain $\mc{X} \subseteq \Rd$, the order-$s$ (Hilbert)-Sobolev space $H^s(\mc{X})$ consists of those functions which are $s$-times weakly differentiable, with order-$1$ up through order-$s$ derivatives bounded in $L^2(\mc{X})$ norm. Equivalently, Sobolev spaces can be defined in a spectral manner;  letting $\psi_1,\psi_2,\ldots$ be the eigenfunctions of a Laplacian operator defined with respect to $\mc{X}$, Sobolev spaces consist of those functions $f$ whose Fourier coefficients $\theta_k := \int_{\mc{X}} f \psi_k$ are bounded in a particular sequence norm (see Section~\ref{subsec:spectral_projection}). The minimax estimation and testing rates for Sobolev spaces are generally speaking well-known---when using mean-squared error as the loss, the estimation rate is $n^{-2s/(2s + d)}$, and the testing rate is $n^{-4s/(4s + d)}$---and are achieved by various methods. However, since Sobolev spaces admit a spectral characterization in terms of $\psi_1,\psi_2,\ldots$, spectral projection methods that use these eigenfunctions as features are a particularly natural choice for regression over Sobolev spaces, and are in a sense canonical estimators for these function classes. 
% (AG 8/13/21): Last clause makes a strong statement that I personally believe, but may irritate certain readers. Keep or remove?
% (AG 8/18/21): I do not love the name ``classical spectral projection method''. It doesn't roll off the tongue, and ``classical'' doesn't really convey what the method is. Is there a better name?

@@

To derive these results, we develop a framework for analysis reminiscent of that  classically used to analyze population-level spectral projection methods, but ultimately quite different. The chief difference is in the approximation error (bias). For our purposes we must establish that with high probability, the eigenvectors of graph Laplacians (which are data-dependent and hence random objects) can be used to effectively approximate continuum Sobolev functions, whereas in the classical analysis an upper bound on the approximation error follows immediately from the aforementioned spectral characterization of Sobolev spaces. To tightly upper bound the approximation error of Laplacian eigenmaps, we rely on several recent results regarding concentration of the spectra of graph Laplacians around their population limits, and prove some additional results of this type which may be of independent interest.

@@

\begin{table}
	\begin{center}
		\begin{tabular}{p{.175\textwidth} p{.175\textwidth} | p{.14\textwidth} p{.12\textwidth} }
			Smoothness order & Dimension & Flat Euclidean (Model~\ref{def:model_flat_euclidean}) & Manifold (Model~\ref{def:model_manifold}) \\
			\hline
			\multirow{2}{*}{$s = 1$} & $\dim(\mc{X}) < 4$ & ${\bf n^{-4s/(4s + d)}}$ & ${\bf n^{-4s/(4s + m)}}$ \\
			& $\dim(\mc{X}) \geq 4$ & ${\bf n^{-1/2}}$ & ${\bf n^{-1/2}}$ \\
			\hline
			\multirow{3}{*}{$s = 2$ or $3$} & $\dim(\mc{X}) \leq 4$  & ${\bf n^{-4s/(4s + d)}}$ & ${\bf n^{-4s/(4s + m)}}$ \\
			& $4 <\dim(\mc{X}) < 4s$  & $n^{-2s/(2(s - 1) + d)}$ & $n^{-2s/(2(s - 1) + m)}$\\
			& $\dim(\mc{X}) \geq 4s$ & ${\bf n^{-1/2}}$ & ${\bf n^{-1/2}}$ \\
			\hline
			\multirow{3}{*}{$s > 3$} & $\dim(\mc{X}) \leq 4$ & ${\bf n^{-4s/(4s + d)}}$ & $n^{-12/(12 + d)}$ \\
			& $4 < \dim(\mc{X}) < 4s$ & $n^{-2s/(2(s - 1) + d)}$ & $n^{-6/(4 + m)}$ \\
			& $\dim(\mc{X}) \geq 4s$ & ${\bf n^{-1/2}}$ & ${\bf n^{-1/2}}$ \\
		\end{tabular}
	\end{center}
	\caption{Summary of Laplacian eigenmaps testing rates over Sobolev balls. Bold font marks minimax optimal rates. Rates when $d > 4s$ assume that $f_0 \in L^4(\mc{X})$, and depend on $\|f_0\|_{L^4(\mc{X})}$. Although we suppress it for simplicity, in all cases when the Laplacian eigenmaps test is optimal, the dependence of the error rate on the radius $M$ of the Sobolev ball is also optimal.}
	\label{tbl:testing_rates}
\end{table}

\subsection{Related work}

@@

\subsection{Roadmap}
We now outline the structure of the rest of this paper. In Section~\ref{sec:setup_main_results}, we give our formal modeling assumptions, and precisely define the Laplacian eigenmaps methods we study. Propositions~\ref{prop:spectral_series_estimation} and ~\ref{prop:spectral_series_testing}, in Section~\ref{subsec:spectral_projection}, show that under rather general (nonparametric) conditions on the design distribution, population-level spectral projection methods achieve minimax rates of convergence over Sobolev classes. Then in Sections~\ref{sec:minimax_optimal_laplacian_eigenmaps},~\ref{sec:manifold_adaptivity}, and \ref{sec:out_of_sample}, we give our main upper bounds on the error of Laplacian eigenmaps (summarized above); these upper bounds hold under similarly general conditions, and imply that Laplacian eigenmaps estimators and tests are also minimax rate-optimal. In Section~\ref{sec:experiments} we examine the empirical behavior of Laplacian eigenmaps, and show that even at moderate sample sizes Laplacian eigenmaps is competitive with population-level spectral projection. We conclude with some discussion in Section~\ref{sec:discussion}.

\subsection{Notation}
We frequently refer to various classical function classes. For an open set $\mc{X}$ equipped with a volume form $d\mu$, we let $L^2(\mc{X})$ denote the set of functions $f$ for which $\|f\|_{L^2(\mc{X})}^2 := \int f^2 \,d\mu  < \infty$, and equip $L^2(\mc{X})$ with the norm $\|\cdot\|_{L^2(\mc{X})}$. We define $\dotp{f}{g}_P := \int fg\,dP$, and let $L^2(P)$ contain those functions $f$ for which $\|f\|_P^2 := \dotp{f}{f}_P$ is finite. Equivalently, we let $L^2(P_n)$ consist of those ``functions'' $f: \{X_1,\ldots,X_n\} \to \Reals$ for which the empirical norm $\|f\|_{n}^2 := \frac{1}{n}\sum_{i = 1}^{n} \bigl(f(X_i)\bigr)^2 < \infty$. When there is no chance of confusion, we will sometimes associate functions in $L^2(P_n)$ with vectors in $\Reals^n$, and vice versa. We use $C^k(\mc{X})$ to refer to functions which are $k$ times continuously differentiable in $\mc{X}$, either for some integer $k \geq 1$ or for $k = \infty$. We let $C_c^{\infty}(\mc{X})$ represent those functions in $C^{\infty}(\mc{X})$ with support $V$ compactly contained in $\mc{X}$, meaning $\wb{V}$ is compact and $\wb{V} \subseteq \mc{X}$. We write $\partial f/\partial r_i$ for the partial derivative of $f$ in the $i$th standard coordinate of $\Rd$, and use the multi-index notation $D^{\alpha}f := \partial^{|\alpha|}f/\partial^{\alpha_1}x_1\ldots\partial^{\alpha_d}x_d$ for multi-indices $\alpha \in \Reals^d$. Recall that for a given multi-index $\alpha \in \mathbb{N}^d$, a function $f$ is \emph{$\alpha$-weakly differentiable} if there exists some $h \in L^1(\mc{X})$ such that
\begin{equation*}
\int_{\mc{X}} h g = (-1)^{|\alpha|} \int_{\mc{X}} f D^{\alpha}g, \quad \textrm{for every $g \in C_c^{\infty}(\mc{X})$.}
\end{equation*}
If such a function $h$ exists, it is the $\alpha$th weak partial derivative of $f$, and denoted by $D^{\alpha}f := h$. For functions $f$ which are $|\alpha|$-times classically differentiable, this coincides with the classical definition of derivative, and so we use the same notation for both.

@@

\subsubsection{Flat Euclidean model}
In Definitions~\ref{def:model_flat_euclidean}-\ref{def:zero_trace_sobolev_space}, we collect the assumptions we make when working under the flat Euclidean model. We begin by giving some regularity conditions on the design.

@@

\begin{definition}[Zero-trace Sobolev space]
	\label{def:zero_trace_sobolev_space}
	A function $f \in H^s(\mc{X})$ belongs to the zero-trace Sobolev space $H_0^s(\mc{X})$ if there exists a sequence $f_1,f_2,\ldots$ of functions in $C_c^{\infty}(\mc{X})$ such that
	\begin{equation*}
	\lim_{k \to \infty}\|f_k - f\|_{H^s(\mc{X})} = 0.
	\end{equation*}
	The normed ball $H_0^{s}(\mc{X};M) := H_0^{s}(\mc{X}) \cap H^{s}(\mc{X};M)$.
\end{definition}
Boundary conditions play an important role in the analysis of spectral methods, as we explain further in Section~\ref{subsec:spectral_projection}. For now, we limit ourselves to pointing out that for functions $f \in C^\infty(\mc{X})$, the zero-trace condition can be stated more concretely, as implying that $\partial^{k}f/\partial{\bf n}^k(x) = 0$ for each $k = 0,\ldots,s - 1$, and for all $x \in \partial\mc{X}$. (Here $\partial/(\partial {\bf n})$ is the partial derivative operator in the direction of the normal vector $\mathbf{n}$.)

\subsubsection{Manifold model}
As in the flat Euclidean case, we start with some regularity conditions on the design. One such condition will be on the reach $R$ of the manifold $\mc{X}$, which we recall is defined as follows:
\begin{equation*}
R := \Bigl\{\sup_{r > 0}: \forall z \in \Rd, \inf_{x \in \mc{X}} \|z - x\| \leq r, ~\exists! y \in \mc{X}~\mathrm{s.t.}~\|z - y\| = \inf_{x \in \mc{X}} \|z - x\|\Bigr\}.
\end{equation*}
In words, the reach is the largest radius of a ball which can be rolled around the manifold $\mc{X}$.
\begin{definition}[Manifold model]
	\label{def:model_manifold}
	The support $\mc{X}$ of the design distribution $P$ is a closed, connected, and smooth Riemannian manifold (without boundary) embedded in $\Rd$, of intrinsic dimension $1 \leq m < d$, and with a positive reach $R > 0$. The design distribution $P$ admits a Lipschitz density $p$ with respect to the volume form $d\mu$ induced by the Riemannian structure of $\mc{X}$, which is bounded away from $0$ and $\infty$,
	\begin{equation*}
	0 < p_{\min} \leq p(x) \leq p_{\max} < \infty, \quad \textrm{for all $x \in \mc{X}$.}
	\end{equation*}
\end{definition}

@@

\subsubsection{Estimation}
We begin with an upper bound on the $L^2(P)$ risk of $\wt{f}$.
\begin{proposition}
	\label{prop:spectral_series_estimation}
	Suppose data is observed according to Model~\ref{def:model_flat_euclidean}, and that the density $p$ is known.  Suppose additionally that $\partial \mc{X} \in C^{1,1}$, $p \in C^{\infty}(\mc{X})$, $f_0 \in \mc{H}^{s}(\mc{X};M)$ and $\|f_0\|_P^2 \leq 1$. Then there exists a constant $C$ which does not depend on $f_0$, $M$ or $n$ such that the following statement holds: if the spectral projection estimator $\wt{f}$ is computed with parameter $K = \max\{\floor{M^2n}^{d/(2s + d)},1\}$, then
	\begin{equation}
	\label{eqn:spectral_series_estimation}
	\Ebb\bigl[\|\wt{f} - f_0\|_P^2\bigr] \leq C \min\Bigl\{M^2\bigl(M^2n\bigr)^{-2s/(2s + d)}, \frac{1}{n}\Bigr\}.
	\end{equation}
\end{proposition}
When the Sobolev ball radius $n^{-1/2} \lesssim M$, the upper bound in~\eqref{eqn:spectral_series_estimation} is on the order of $M^2(M^2n)^{-2s/(2s + d)}$, which is the well-known minimax rate for estimation over Sobolev classes (see e.g.~\cite{gyorfi2006,wasserman2006,tsybakov08} and references therein, and specifically Theorem~3.2 of~\cite{gyorfi2006} for a matching lower bound in the context of nonparametric regression with random design).

@@

On the other hand, when $s \leq d/4$ the minimax regression testing rates over $H^s(\mc{X})$ are not known. If one explicitly assumes $f_0 \in L^4(\mc{X};1)$---note that $H^s(\mc{X})$ does not continuously embed into $L^4(\mc{X})$ when $s \leq d/4$---then the minimax critical radius for regression testing is on the order of $n^{-1/2}$ \citep{guerre02}, and is achieved by a test using the naive statistic $\|{\bf Y}\|_n^2$. In other words, the regression testing problem over Sobolev spaces fundamentally changes when $s \leq d/4$, and hereafter when we discuss testing we will limit our consideration to $s > d/4$. 
\begin{proof}{Proof of Proposition~\ref{prop:spectral_series_testing}.}
We briefly lay out the main ideas needed to prove Proposition~\ref{prop:spectral_series_testing}, following the lead of~\cite{ingster2009} who prove a similar result in the special case where $M = 1$ and $P$ is the uniform distribution over $\mc{X} = [0,1]^d$, and referring to that work for more details.

>>>>

\section{Introduction}
\label{sec:introduction}
Two fundamental statistical goals are to learn and leverage structure of an unknown distribution on the basis of samples. Neighborhood graph Laplacians, and their associated spectra, have proven to be powerful and general tools for both purposes. They have been successfully used for various statistical tasks such as (spectral) clustering, manifold learning, level-set estimation, semi-supervised learning, etc.

@@

More specifically, we consider the worst-case risk of Laplacian eigenmaps over normed balls in Sobolev spaces. Roughly speaking, for a domain $\mc{X} \subseteq \Rd$, the order-$s$ (Hilbert-)Sobolev space $H^s(\mc{X})$ consists of those functions which are $s$-times weakly differentiable, with order-$1$ up through order-$s$ derivatives bounded in $L^2(\mc{X})$ norm. Equivalently, Sobolev spaces can be defined in a spectral manner;  letting $\psi_1,\psi_2,\ldots$ be the eigenfunctions of a Laplacian operator defined with respect to $\mc{X}$, Sobolev spaces consist of those functions $f$ whose Fourier coefficients $\theta_k := \int_{\mc{X}} f \psi_k$ are bounded in a particular sequence norm (see Section~\ref{subsec:spectral_projection}). The minimax estimation and testing rates for Sobolev spaces are generally speaking well-known---when using mean-squared error as the loss, the estimation rate is $n^{-2s/(2s + d)}$, and the testing rate is $n^{-4s/(4s + d)}$---and are achieved by various methods. However, since Sobolev spaces admit a spectral characterization in terms of $\psi_1,\psi_2,\ldots$, spectral projection methods that use these eigenfunctions as features are a particularly natural choice for regression over Sobolev spaces, and are in a sense canonical estimators for these function classes. 
% (AG 8/13/21): Last clause makes a strong statement that I personally believe, but may irritate certain readers. Keep or remove?
% (AG 8/18/21): I do not love the name ``classical spectral projection method''. It doesn't roll off the tongue, and ``classical'' doesn't really convey what the method is. Is there a better name?

@@

To derive these results, we develop a framework for analysis reminiscent of that  classically used to analyze population-level spectral projection estimators, but ultimately quite different. The chief difference is in the approximation error (bias). For our purposes we must establish that with high probability, the eigenvectors of graph Laplacians (which are data-dependent and hence random objects) can be used to effectively approximate continuum Sobolev functions, whereas in the classical analysis an upper bound on the approximation error follows immediately from the aforementioned spectral characterization of Sobolev spaces. To tightly upper bound the approximation error of Laplacian eigenmaps, we rely on several recent results regarding concentration of the spectra of graph Laplacians around their population limits, and prove some additional results of this type which may be of independent interest.

@@

\begin{table}
	\begin{center}
		\begin{tabular}{p{.175\textwidth} p{.175\textwidth} | p{.14\textwidth} p{.12\textwidth} }
			Smoothness order & Dimension & Flat Euclidean (Model~\ref{def:model_flat_euclidean}) & Manifold (Model~\ref{def:model_manifold}) \\
			\hline
			\multirow{2}{*}{$s = 1$} & $\dim(\mc{X}) < 4$ & ${\bf n^{-4s/(4s + d)}}$ & ${\bf n^{-4s/(4s + m)}}$ \\
			& $\dim(\mc{X}) \geq 4$ & ${\bf n^{-1/2}}$ & ${\bf n^{-1/2}}$ \\
			\hline
			\multirow{3}{*}{$s = 2$ or $3$} & $\dim(\mc{X}) \leq 4$  & ${\bf n^{-4s/(4s + d)}}$ & ${\bf n^{-4s/(4s + m)}}$ \\
			& $4 <\dim(\mc{X}) < 4s$  & $n^{-2s/(2(s - 1) + d)}$ & $n^{-2s/(2(s - 1) + m)}$\\
			& $\dim(\mc{X}) \geq 4s$ & ${\bf n^{-1/2}}$ & ${\bf n^{-1/2}}$ \\
			\hline
			\multirow{3}{*}{$s > 3$} & $\dim(\mc{X}) \leq 4$ & ${\bf n^{-4s/(4s + d)}}$ & $n^{-12/(12 + d)}$ \\
			& $4 < \dim(\mc{X}) < 4s$ & $n^{-2s/(2(s - 1) + d)}$ & $n^{-6/(4 + m)}$ \\
			& $\dim(\mc{X}) \geq 4s$ & ${\bf n^{-1/2}}$ & ${\bf n^{-1/2}}$ \\
		\end{tabular}
	\end{center}
	\caption{Summary of Laplacian eigenmaps testing rates over Sobolev balls. Bold font marks minimax optimal rates. Rates when $d > 4s$ assume that $f_0 \in L^4(\mc{X})$, and depend on $\|f_0\|_{L^4(\mc{X})}$. Although we suppress it for simplicity, in all cases when othe Laplacian eigenmaps test is optimal, the dependence of the error rate on the radius $M$ of the Sobolev ball is also optimal.}
	\label{tbl:testing_rates}
\end{table}

\paragraph{Related work.}

@@

\paragraph{Roadmap.}
We now outline the structure of the rest of this paper. In Section~\ref{sec:setup_main_results}, we give our formal modeling assumptions, and precisely define the Laplacian eigenmaps methods we study. Propositions~\ref{prop:spectral_series_estimation} and ~\ref{prop:spectral_series_testing}, in Section~\ref{subsec:spectral_projection}, show that under rather general (nonparametric) conditions on the design distribution, population-level spectral projection methods achieve minimax rates of convergence over Sobolev classes. Then in Sections~\ref{sec:minimax_optimal_laplacian_eigenmaps},~\ref{sec:manifold_adaptivity}, and \ref{sec:out_of_sample}, we give our main upper bounds on the error of Laplacian eigenmaps (summarized above); these upper bounds hold under similarly general conditions, and imply that Laplacian eigenmaps estimators and tests are also minimax rate-optimal. In Section~\ref{sec:experiments} we examine the empirical behavior of Laplacian eigenmaps, and show that even at moderate sample sizes Laplacian eigenmaps is competitive with population-level spectral projection. We conclude with some discussion in Section~\ref{sec:discussion}.

\paragraph{Notation.}
We frequently refer to various classical function classes. For an open set $\mc{X}$ equipped with a volume form $d\mu$, we let $L^2(\mc{X})$ denote the set of functions $f$ for which $\|f\|_{L^2(\mc{X})}^2 := \int f^2 \,d\mu  < \infty$, and equip $L^2(\mc{X})$ with the norm $\|\cdot\|_{L^2(\mc{X})}$. We define $\dotp{f}{g}_P := \int fg\,dP$, and let $L^2(P)$ contain those functions $f$ for which $\|f\|_P^2 := \dotp{f}{f}_P$ is finite. Equivalently, we let $L^2(P_n)$ consist of those ``functions'' $f: \{X_1,\ldots,X_n\} \to \Reals$ for which the empirical norm $\|f\|_{n}^2 := \frac{1}{n}\sum_{i = 1}^{n} \bigl(f(X_i)\bigr)^2 < \infty$. When there is no chance of confusion, we will sometimes associate functions in $L^2(P_n)$ with vectors in $\Reals^n$, and vice versa. We use $C^k(\mc{X})$ to refer to functions which are $k$ times continuously differentiable in $\mc{X}$, either for some integer $k \geq 1$ or for $k = \infty$. We let $C_c^{\infty}(\mc{X})$ represent those functions in $C^{\infty}(\mc{X})$ with support $V$ compactly contained in $\mc{X}$, meaning $\wb{V}$ is compact and $\wb{V} \subseteq \mc{X}$. We write $\partial f/\partial r_i$ for the partial derivative of $f$ in the $i$th standard coordinate of $\Rd$, and use the multi-index notation $D^{\alpha}f := \partial^{|\alpha|}f/\partial^{\alpha_1}x_1\ldots\partial^{\alpha_d}x_d$ for multi-indices $\alpha \in \Reals^d$. Recall that for a given multi-index $\alpha \in \mathbb{N}^d$, a function $f$ is \emph{$\alpha$-weakly differentiable} if there exists some $h \in L^1(\mc{X})$ such that
\begin{equation*}
\int_{\mc{X}} h g = (-1)^{|\alpha|} \int_{\mc{X}} f D^{\alpha}g, \quad \textrm{for every $g \in C_c^{\infty}(\mc{X})$.}
\end{equation*}
If such a function $h$ exists, it is the $\alpha$th weak partial derivative of $f$, and denoted by $D^{\alpha}f := h$. For functions $f$ which are $|\alpha|$-times classically differentiable, this coincides with the classical definition of derivative, and so we use the same notation for both.

@@

\paragraph{Flat Euclidean model.}
In Definitions~\ref{def:model_flat_euclidean}-\ref{def:zero_trace_sobolev_space}, we collect the assumptions we make when working under the flat Euclidean model. We begin by giving some regularity conditions on the design.

@@

\begin{definition}[Zero-trace Sobolev space]
	\label{def:zero_trace_sobolev_space}
	A function $f \in H^s(\mc{X})$ belongs to the zero-trace Sobolev space $H_0^s(\mc{X})$ if there exists a sequence $f_1,f_2,\ldots$ of functions in $C_c^{\infty}(\mc{X})$ such that
	\begin{equation*}
	\lim_{k \to \infty}\|f_k - f\|_{H^s(\mc{X})} = 0.
	\end{equation*}
	The normed ball $H_0^{s}(\mc{X};M) := H_0^{s}(\mc{X}) \cap H^{s}(\mc{X};M)$.
\end{definition}
Boundary conditions play an important role in the analysis of spectral methods, as we explain further in Section~\ref{subsec:spectral_projection}. For now, we limit ourselves to pointing out that for functions $f \in C^\infty(\mc{X})$, the zero-trace condition can be stated more concretely, as implying that $\partial^{k}f/\partial{\bf n}^k(x) = 0$ for each $k = 0,\ldots,s - 1$, and for all $x \in \partial\mc{X}$. (Here $\partial/(\partial {\bf n})$ is the partial derivative operator in the direction of the normal vector $\mathbf{n}$.)
\paragraph{Manifold model.}
As in the flat Euclidean case, we start with some regularity conditions on the design. One such condition will be on the reach $R$ of the manifold $\mc{X}$, which we recall is defined as follows:
\begin{equation*}
R := \Bigl\{\sup_{r > 0}: \forall z \in \Rd, \inf_{x \in \mc{X}} \|z - x\| \leq r, ~\exists! y \in \mc{X}~\mathrm{s.t.}~\|z - y\| = \inf_{x \in \mc{X}} \|z - x\|\Bigr\}.
\end{equation*}
In words, the reach is the largest radius of a ball which can be rolled around the manifold $\mc{X}$.
\begin{definition}[Manifold model]
	\label{def:model_manifold}
	The support $\mc{X}$ of the design distribution $P$ is a closed, connected, and smooth Riemannian manifold (without boundary) embedded in $\Rd$, of intrinsic dimension $1 \leq m < d$, and with a positive reach $R > 0$. The design distribution $P$ admits a Lipschitz density $p$ with respect to the volume form $d\mu$ induced by the Riemannian structure of $\mc{X}$, which is bounded away from $0$ and $\infty$,
	\begin{equation*}
	0 < p_{\min} \leq p(x) \leq p_{\max} < \infty, \quad \textrm{for all $x \in \mc{X}$.}
	\end{equation*}
\end{definition}

@@

\paragraph{Estimation.}
We begin with an upper bound on the $L^2(P)$ risk of $\wt{f}$.
\begin{proposition}
	\label{prop:spectral_series_estimation}
	Suppose data is observed according to Model~\ref{def:model_flat_euclidean}, and that the density $p$ is known.  Suppose additionally that $\partial \mc{X} \in C^{1,1}$, $p \in C^{\infty}(\mc{X})$, $f_0 \in \mc{H}^{s}(\mc{X};M)$ and $\|f_0\|_P^2 \leq 1$. Then there exists a constant $C$ which does not depend on $f_0$, $M$ or $n$ such that the following statement holds: if the spectral projection estimator $\wt{f}$ is computed with parameter $K = \max\{\floor{M^2n}^{d/(2s + d)},1\}$, then
	\begin{equation}
	\label{eqn:spectral_series_estimation}
	\Ebb\bigl[\|\wt{f} - f_0\|_P^2\bigr] \leq C \min\Bigl\{M^2\bigl(M^2n\bigr)^{-2s/(2s + d)}, \frac{1}{n}\Bigr\}.
	\end{equation}
\end{proposition}
When the Sobolev ball radius $n^{-1/2} \lesssim M$, the upper bound in~\eqref{eqn:spectral_series_estimation} is on the order of $M^2(M^2n)^{-2s/(2s + d)}$, which is the well-known minimax rate for estimation over Sobolev classes (see e.g.~\cite{gyorfi2006,wasserman2006,tsybakov08} and references therein, and specifically Theorem~3.2 of~\cite{gyorfi2006} for a matching lower bound in the context of nonparametric regression with random design).

@@

On the other hand, when $s \leq d/4$ the minimax regression testing rates over $H^s(\mc{X})$ are not known. If one explicitly assumes $f_0 \in L^4(\mc{X};1)$---note that $H^s(\mc{X})$ does not continuously embed into $L^4(\mc{X})$ when $s \leq d/4$---then the minimax critical radius for regression testing is on the order of $n^{-1/2}$ \citep{guerre02}, and is achieved by a test using the naive statistic $\|{\bf Y}\|_n^2$. In other words, the regression testing problem over Sobolev spaces fundamentally changes when $s \leq d/4$, and hereafter when we discuss testing we will limit our consideration to $s > d/4$. 
\paragraph{Proof of Proposition~\ref{prop:spectral_series_testing}.}
We briefly lay out the main ideas needed to prove Proposition~\ref{prop:spectral_series_testing}, following the lead of~\cite{ingster2009} who prove a similar result in the special case where $M = 1$ and $P$ is the uniform distribution over $\mc{X} = [0,1]^d$, and referring to that work for more details.
